Simple repression draft

### 2020-06-05 doc/appendix_02_bursty.tex
Not quite final pass on everything, Appdx S3 still needs more

In order to solve Eq.~\ref{eq:si_master_unreg} we will use the generating
function method \mrm{reference?}. The probability generating function is defined
as
\begin{align}
F(z,t) = \sum_{m=0}^\infty z^m p(m,t),
\end{align}
where $z$ is just a dummy variable that will help us later on to obtain the
moments of the distribution. Let us now multiply both sides of
Eq.~\ref{eq:si_master_unreg} by $z^m$ and sum over all $m$
\begin{equation}
\sum_m z^m {d \over d\tau} p(m, \tau) = 
\sum_m z^m \left[ 
- m p(m, \tau) 
+ (m + 1) p(m + 1, \tau) 
+ \lambda \sum_{m' = 0}^m G_{m-m'} p(m', \tau) - \lambda p(m, \tau)
\right],
\end{equation}
where we use $\sum_m \equiv \sum_{m=0}^\infty$. We can distribute the sum and
use the definition of $F(z, t)$ to obtain
\begin{equation}
{d F(z, \tau) \over d\tau} =
- \sum_m z^m m p(m, \tau) +
+ \sum_m z^m (m + 1) p(m + 1, \tau)
+ \lambda \sum_m z^m \sum_{m'=0}^m G_{m-m'} p(m', \tau)
- \lambda F(z, \tau).
\label{eq:si_generating_01}
\end{equation}
We can make use of properties of the generating function to write everything in
terms of $F(z, \tau)$: the first term on the right hand side of
Eq.~\ref{eq:si_generating_01} can be rewritten as
\begin{align}
\sum_{m} z^{m} \cdot m \cdot p(m, \tau) &=
\sum_{m} z \frac{\partial z^{m}}{\partial z} p(m, \tau), \\
&=\sum_{m} z \frac{\partial}{\partial z}\left(z^{m} p(m, \tau)\right), \\
&=z \frac{\partial}{\partial z}\left(\sum_{m} z^{m} p(m, \tau)\right), \\
&= z {\partial F(z, \tau) \over \partial z}.
\end{align}
For the second term on the right hand side of Eq.~\ref{eq:si_generating_01} we
define $k \equiv m + 1$. This allows us to write
\begin{align}
\sum_{m=0}^{\infty} z^{m} \cdot(m+1) \cdot p(m+1, \tau) &=
\sum_{k=1}^{\infty} z^{k-1} \cdot k \cdot p(k, \tau), \\
&=z^{-1} \sum_{k=1}^{\infty} z^{k} \cdot k \cdot p(k, \tau), \\
&=z^{-1} \sum_{k=0}^{\infty} z^{k} \cdot k \cdot p(k, \tau), \\
&=z^{-1} \left(z \frac{\partial F(z)}{\partial z}\right), \\
&=\frac{\partial F(z)}{\partial z}.
\end{align}

@@

The next step requires us to look slightly ahead into what we expect to obtain.
We are working on deriving an equation for the generating function $F(z, \tau)$
that when solved will allow us to compute what we care about, i.e. the
probability function $p(m, \tau)$. Upon finding the function for $F(z, \tau)$,
we will recover this probability distribution by evaluating $F(z=1, \tau)$ in
other cases we can evaluate $F(z=0, \tau)$ to recover instead the moments of the
distribution. The point here is that we know that when all the dust settles we
will evaluate $z$ to be less than one \mrm{someone needs to fact check me}.
Furthermore, we know that the parameter of the geometric distribution $\theta$
must be between zero and one. With these two facts we can safely state that $| z
(1 - \theta) | < 1$. Defining $n \equiv m - m'$ we rewrite the last sum in
Eq.~\ref{eq:si_generating_02} as
\begin{align}
\sum_{m=m'}^{\infty} \left[z (1-\theta)\right]^{m} &= 
\sum_{n=0}^{\infty} \left[z (1-\theta)\right]^{n + m'} \\
&= \left[ z (1 - \theta) \right]^{m'} 
\sum_{n=0}^{\infty} \left[ z (1 - \theta) \right]^{n} \\
&= \left[ z (1 - \theta) \right]^{m'} 
\left( {1 \over 1 - z (1 - \theta)} \right),
\end{align}
where we use the geometric series since, as stated before, $| z (1 - \theta) | <
1$. Putting these results together, the PDE for the generation function is
\begin{equation}
{\partial F \over \partial \tau} = 
{\partial F \over\partial z}
- z {\partial F \over \partial z} - \lambda F
+ \frac{\lambda\theta F}{1-z(1-\theta)}.
\end{equation}
Changing variables to $\xi=1-\theta$ and simplifying gives
\begin{align}
{\partial F \over \partial \tau} + (z - 1) {\partial F \over \partial z} = 
\frac{(z-1)\xi}{1-z\xi}\lambda F.
\label{eq:1state_unreg_015}
\end{align}

@@

To get at the mRNA distribution at steady state we first must solve
Eq.~\ref{eq:1state_unreg_015} setting the time derivative to zero. At
steady-state, the PDE reduces to the ODE
\begin{align}
\deriv[F]{z} = \frac{\xi}{1-z\xi}\lambda F,
\end{align}
which we can integrate as
\begin{align}
\int \frac{dF}{F} = \int \frac{\lambda\xi dz}{1-\xi z}.
\end{align}
The initial conditions for generating functions can be subtle and confusing. The
key fact follows from the definition
$F(z,t) = \sum_m z^m p(m,t)$,
normalization requires that
$F(z=1, t) = \sum_m p(m,t) = 1$.
\footnote{
Sometimes the generating function may be undefined \textit{at} $z=1$ but the
limit still holds. Also many authors change variables from $z$ to other
things, so one must keep track of how the normalization condition transforms.
}
Doing the integrals (and producing constant $c$) gives
\begin{align}
\ln F &= -\lambda \ln(1-\xi z) + c
\\
F &= \frac{c}{(1-\xi z)^\lambda}.
\end{align}
Only one choice for $c$ can satisfy initial conditions, producing
\begin{align}
F(z) = \left(\frac{1-\xi}{1-\xi z}\right)^\lambda
        = \left(\frac{\theta}{1 - z(1-\theta)}\right)^\lambda,
\label{eq:gen_fn}
\end{align}

@@

Let us move from a one-state promoter to a two-state promoter, where one state
has repressor bound and the other produces transcriptional bursts as above.
A schematic of this model is shown as model 5 in
Figure~\ref{fig1:means_cartoons}(C). Although now we have an equation for each
promoter state, otherwise the master equation reads similarly to the one-state
case, except with additional terms corresponding to transitions between promoter
states, namely
\begin{align}
{d\over dt}p_R(m,t) =& 
k_R^+ p_A(m,t) - k_R^- p_R(m,t)
        + (m+1)\gamma p_R(m+1,t) - m\gamma p_R(m,t)
\\
\begin{split}
{d\over dt}p_A(m,t) =& - k_R^+ p_A(m,t) + k_R^- p_R(m,t)
        + (m+1)\gamma p_A(m+1,t) - m\gamma p_A(m,t) 
\\
&- k_i p_A(m,t) + k_i \sum_{m^\prime=0}^m \theta(1-\theta)^{m-m^\prime} p_A(m^\prime,t),
\end{split}
\end{align}
where $p_R(m,t)$ is the probability of the system having $m$ mRNA copies and
having repressor bound to the promoter at time $t$, and $p_A$ is an analogous
probability to find the promoter without repressor bound. $k_R+$ and $k_R^-$
are, respectively, the rates at which repressors bind and unbind to and from the
promoter, and $\gamma$ is the mRNA degradation rate. $k_i$ is the rate at which
bursts initiate, and as before, the geometric distribution of burst sizes has
mean $b=(1-\theta)/\theta$.
\footnote{
$\theta$ can be viewed as a Bernoulli trial probability: after each transcript
in a burst is produced, $\theta$ is the probability that the burst terminates
and $(1-\theta)$ is the probability that the burst continues.
}

@@

\mrm{the use of variable $\gamma$ here is confusing. Is it okay to change it to
$\delta$ or something else in order to avoid associating it with the mRNA
degradation rate?}

@@

f_R(w) = C_1 {_2F_1}(\alpha, \beta, \gamma; w)
+ C_2 w^{1-\gamma}{_2F_1}(1+\alpha-\gamma, 1+\beta-\gamma, 2-\gamma; w)

@@

\gamma &= 1 + k_R^- + k_R^+

@@

and constants $C_1$ and $C_2$ to be set by boundary conditions. Solving for

@@

\footnote{

@@

\gamma &= 1 + k_R^- + k_R^+.

@@

we can immediately set $C_2=0$ to discard the second solution.

@@

so $\gamma>1$ and therefore $w^{1-\gamma}$ blows up as $w\rightarrow0$.

@@

Plugging our solution for $f_R(w) = C_1{_2F_1}(\alpha, \beta, \gamma; w)$

@@

\pderiv[f_R]{w} = C_1\frac{\alpha\beta}{\gamma}
                {_2F_1}(\alpha+1, \beta+1, \gamma+1; w),

@@

f_A(w) = \frac{C_1}{k_R^+}

@@

k_R^- {_2F_1}(\alpha, \beta, \gamma; w)
+ w\frac{\alpha\beta}{\gamma} {_2F_1}(\alpha+1, \beta+1, \gamma+1; w)

@@

f(w) = C_1\left(1 + \frac{k_R^-}{k_R^+}\right)
        {_2F_1}(\alpha, \beta, \gamma; w)
+ w \frac{C_1}{k_R^+} \frac{\alpha\beta}{\gamma}
        {_2F_1}(\alpha+1, \beta+1, \gamma+1; w).

@@

w\frac{\alpha\beta}{\gamma} {_2F_1}(\alpha+1, \beta+1, \gamma+1; w)

@@

(\gamma-1)\left(
{_2F_1}(\alpha, \beta, \gamma-1; w) - {_2F_1}(\alpha, \beta, \gamma; w)

@@

f(w) = \frac{C_1}{k_R^+}

@@

        {_2F_1}(\alpha, \beta, \gamma; w)
+ (\gamma-1)\left(
        {_2F_1}(\alpha, \beta, \gamma-1; w) - {_2F_1}(\alpha, \beta, \gamma; w)

@@

and since $\gamma-1 = k_R^+ + k_R^-$, the first and third terms cancel,

@@

f(w) = C_1\frac{k_R^+ + k_R^-}{k_R^+} {_2F_1}(\alpha, \beta, \gamma-1; w).

@@

${_2F_1}(\alpha, \beta, \gamma-1; 0) = 1$, so we must have
$C_1 = k_R^+ / (k_R^+ + k_R^-)$ and consequently

@@

$p(m)$ and similarly for $\beta$ and $\gamma$. These resemble ratios of
factorials, but since $\alpha$, $\beta$, and $\gamma$ are not necessarily

@@

\subsubsection{Moments of two-state bursty promoter mRNA distribution}

We can verify the correctness of~\eq{eq:p_m_bursty+rep_appdx} by comparison with
Gillespie simulation. We can also easily compute moments of the distribution
from the generating function, and check these against raw moments computed
directly from the master equation.

The $j$-th raw moment of the distribution can be computed from the generating
function from the formula
\begin{align}
\langle m^j \rangle
= \left. \left(z \frac{\partial}{\partial z}\right)^j f(z)\right|_{z=1}.
\end{align}
Note that \textit{probabilities} are generated by differentiating and evaluating
at $z=0$, while \textit{moments} are generated by differentiating and evaluating
at $z=1$. This also aligns with the normalization constraint, which can be
thought of as the zeroth raw moment.

The mean mRNA is given by
\begin{align}
\langle m \rangle = \left.z \pderiv[f]{z} \right|_{z=1}
= \frac{\alpha\beta b}{k_R^+ + k_R^-}
= k_i b\frac{k_R^-}{k_R^+ + k_R^-},
\end{align}
which is intuitive and appealingly simple: the mean mRNA is just the mean mRNA
in the absence of repression ($k_i b$, the burst rate time mean burst size)
times the fraction of time the repressor is bound, $k_R^-/(k_R^+ + k_R^-)$.

To compute the variance, let us first compute
\begin{align}
\begin{split}
\langle m^2\rangle
&= \left .z\pderiv{z}\left(z\pderiv[f]{z}\right) \right|_{z=1}
= \left. z\left(\pderiv[f]{z} + z\psecderiv[f]{z}\right)\right|_{z=1}
\\
&= \frac{\alpha\beta b}{k_R^+ + k_R^-}
        + \frac{\alpha(\alpha+1)\beta(\beta+1) b^2}
                {(k_R^+ + k_R^-)(k_R^+ + k_R^- + 1)}
\\
&= \frac{k_i b k_R^-}{k_R^+ + k_R^-}
        + \frac{k_i k_R^- b^2 (1 + k_i + k_R^+ + k_R^- + k_i k_R^-)}
                {(k_R^+ + k_R^-)(k_R^+ + k_R^- + 1)},
\end{split}
\end{align}
from which the variance is
\begin{align}
\begin{split}
var(m) = \langle m^2\rangle - \langle m\rangle^2
&= \frac{k_i b k_R^-}{k_R^+ + k_R^-}
\left(
1 + \frac{b(1 + k_i + k_R^+ + k_R^- + k_i k_R^-)} {k_R^+ + k_R^- + 1}
- \frac{k_i b k_R^-}{k_R^+ + k_R^-}
\right)
\\
&= k_i b\frac{k_R^-}{k_R^+ + k_R^-}
\left(
1 + b + k_i b\frac{1 + k_R^-} {k_R^+ + k_R^- + 1}
- \frac{k_i b k_R^-}{k_R^+ + k_R^-}
\right)
\\
&= k_i b\frac{k_R^-}{k_R^+ + k_R^-}
\left(
1 + b + k_i b\frac{k_R^+}{(k_R^+ + k_R^-)(k_R^+ + k_R^- + 1)}
\right).
\end{split}
\end{align}
The Fano factor is the quantity in parentheses. This can be interpreted as
the variability arising from the negative binomial distribution of the
unregulated promoter $(1+b)$ plus additional noise arising from repression.

\mmnote{Still to writeup: moment derivation direct from CME to compare}
\mrm{stopped reading here}

@@

The hypergeometric function is defined by its Taylor series as
\begin{align}
{_2F_1}(a,b,c;z) 
= \sum_{l=0}^\infty
\frac{\Gamma(a + l)\Gamma(b + l)\Gamma(c)}
        {\Gamma(a)\Gamma(b)\Gamma(c + l)}
\frac{z^l}{l!}
\end{align}
for $|z|<1$, and by analytic continuation elsewhere.
If $z\lesssim1/2$ and $\alpha$ and $\beta$ are not too large
(absolute value below 20 or 30),
then the series converges quickly and an accurate numerical representation is
easily computed by truncating the series after a reasonable number of terms.
Unfortunately, we need to evaluate ${_2F_1}$ over mRNA copy numbers fully out
to the tail of the distribution, which can easily reach 50, possibly 100.
From~\eq{eq:p_m_bursty+rep_appdx}, this means evaluating ${_2F_1}$
repeatedly for values of $a$, $b$, and $c$ spanning the full range
from $\mathcal{O}(1)$ to $\mathcal{O}(10^2)$,
even if $\alpha$, $\beta$, and $\gamma$
in~\eq{eq:p_m_bursty+rep_appdx} are small,
with the situation even worse if they are not small.
A naive numerical evaluation of the series definition will be
prone to overflow and, if any of $a,b,c<0$, then some successive terms in the
series have alternating signs which can lead to catastrophic cancellations.

@@

This provides a convenient solution for our problem: we need repeated
evaluations of ${_2F_1}(a+m, b+m, c+m; z)$
for fixed $a,b$, and $c$ and many integer values of $m$.
They idea is that we can use arbitrary precision arithmetic to evaluate
${_2F_1}$ for just two particular values of $m$ and then generate
${_2F_1}$ for the other 50-100 values of $m$ using the recurrence
relation.\footnote{
There are even more sophisticated ways of utilizing the recurrence
relations that might have netted another factor of 2 speed-up, and
possibly as much as a factor of 10, but the method described here had
already reduced the computation time to an acceptable
$\mathcal{O}(\text{1 min})$, so these more sophisticated approaches did
not seem worth the time to pursue.
}

However, there two further wrinkles.
The first is that
a naive application of the recurrence relation is numerically unstable.
Roughly, this is because the three term recurrence relations,
like second order ODEs, admit two linearly independent solutions.
In a certain eigenbasis, one of these solutions dominates the other
as $n\rightarrow\infty$, and as $n\rightarrow-\infty$,
the dominance is reversed.
If we fail to work in this eigenbasis, our solution of the recurrence relation
will be a mixture of these solutions and rapidly accumulate numerical error.
For our purposes, it suffices to know that the authors of~\cite{Gil2007}
derived the numerically stable solutions (so-called \textit{minimal solutions})
for several possible choices of $\epsilon_i, \epsilon_j, \epsilon_k$.
Running the recurrence in the proper direction using a minimal solution
is numerically robust and can be done entirely in floating point arithmetic, 
so that we only need to evaluate ${_2F_1}$ with arbitrary precision arithmetic
to generate the seed values for the recursion.

@@

\subsubsection{Particulars}
Let us now demonstrate these generalities for our problem of interest.
In order to evaluate the probability distribution of our
model,~\eq{eq:p_m_bursty+rep_appdx}, we need to evaluate hypergeometric functions
of the form ${_2F_1}(\alpha+m, \beta+m, \gamma+m; -b)$
for values of $m$ ranging from $0$ to $\mathcal{O}(100)$.
The authors of~\cite{Gil2007} did not derive a recursion relation
for precisely this case. We could follow their methods and do so ourselves,
but it is much easier to convert to a case that they did consider.
The strategy is to look through the minimal solutions tabulated
in~\cite{Gil2007} and search for a transformation we could apply to
${_2F_1}(\alpha+m, \beta+m, \gamma+m; -b)$ that would place the $m$'s
(the variable being incremented by the recursion)
in the same arguments of ${_2F_1}$ as the minimal solution.
After some ``guess and check,'' we found that the transformation
\begin{align}
{_2F_1}(\alpha+m, \beta+m, \gamma+m; -b)
=
(1+b)^{-\alpha-m}
        {_2F_1}\left(\alpha+m, \gamma-\beta, \gamma+m; \frac{b}{1+b}\right),
\label{eq:rec_euler_pretransform}
\end{align}
produces a ${_2F_1}$ on the right hand side that closely resembles
the minimal solutions $y_{3,m}$ and $y_{4,m}$ in Eq.~4.3 in~\cite{Gil2007}.
Explicitly, these solutions are
\begin{align}
y_{3,m}
&\propto
{_2F_1}\left(-\alpha^\prime + \gamma^\prime - m,
                -\beta^\prime + \gamma^\prime,
                1-\alpha^\prime-\beta^\prime+\gamma^\prime-m;
                1-z\right)
\\
y_{4,m}
&\propto
{_2F_1}\left(\alpha^\prime + m,
                \beta^\prime,
                1+\alpha^\prime+\beta^\prime-\gamma^\prime+m;
                1-z\right),
\label{eq:minimal_soln_sans_prefac}
\end{align}
where we have omitted prefactors which are unimportant for now.
Which of these two we should use depends on what values $z$ takes on.
Equating $1-z=b/(1+b)$ gives $z=1/(1+b)$, and since $b$ is strictly positive,
$z$ is bounded between 0 and 1.
From Eq.~4.5 in~\cite{Gil2007}, $y_{4,m}$ is the minimal solution
for real $z$ satisfying $0<z<2$, so this is the only minimal solution we need.

>>>>

In order to solve Eq.~\ref{eq:si_master_unreg} we will use the generating
function method~\cite{vanKampen2007}. The probability generating function
is defined as
\begin{align}
F(z,t) = \sum_{m=0}^\infty z^m p(m,t),
\end{align}
where $z$ is just a dummy variable that will help us later on to obtain the
moments of the distribution. Let us now multiply both sides of
Eq.~\ref{eq:si_master_unreg} by $z^m$ and sum over all $m$
\begin{equation}
\sum_m z^m {d \over d\tau} p(m, \tau) = 
\sum_m z^m \left[ 
- m p(m, \tau) 
+ (m + 1) p(m + 1, \tau) 
+ \lambda \sum_{m' = 0}^m G_{m-m'} p(m', \tau) - \lambda p(m, \tau)
\right],
\end{equation}
where we use $\sum_m \equiv \sum_{m=0}^\infty$. We can distribute the sum and
use the definition of $F(z, t)$ to obtain
\begin{equation}
{d F(z, \tau) \over d\tau} =
- \sum_m z^m m p(m, \tau)
+ \sum_m z^m (m + 1) p(m + 1, \tau)
+ \lambda \sum_m z^m \sum_{m'=0}^m G_{m-m'} p(m', \tau)
- \lambda F(z, \tau).
\label{eq:si_generating_01}
\end{equation}
We can make use of properties of the generating function to write everything in
terms of $F(z, \tau)$: the first term on the right hand side of
Eq.~\ref{eq:si_generating_01} can be rewritten as
\begin{align}
\sum_{m} z^{m} \cdot m \cdot p(m, \tau) &=
\sum_{m} z \frac{\partial z^{m}}{\partial z} p(m, \tau), \\
&=\sum_{m} z \frac{\partial}{\partial z}\left(z^{m} p(m, \tau)\right), \\
&=z \frac{\partial}{\partial z}\left(\sum_{m} z^{m} p(m, \tau)\right), \\
&= z {\partial F(z, \tau) \over \partial z}.
\end{align}
For the second term on the right hand side of Eq.~\ref{eq:si_generating_01} we
define $k \equiv m + 1$. This allows us to write
\begin{align}
\sum_{m=0}^{\infty} z^{m} \cdot(m+1) \cdot p(m+1, \tau) &=
\sum_{k=1}^{\infty} z^{k-1} \cdot k \cdot p(k, \tau), \\
&=z^{-1} \sum_{k=1}^{\infty} z^{k} \cdot k \cdot p(k, \tau), \\
&=z^{-1} \sum_{k=0}^{\infty} z^{k} \cdot k \cdot p(k, \tau), \\
&=z^{-1} \left(z \frac{\partial F(z)}{\partial z}\right), \\
&=\frac{\partial F(z)}{\partial z}.
\end{align}

@@

The next step requires us to look slightly ahead into what we expect to obtain.
We are working on deriving an equation for the generating function $F(z, \tau)$
that when solved will allow us to compute what we care about, i.e. the
probability function $p(m, \tau)$. Upon finding the function for $F(z, \tau)$,
we will recover this probability distribution by evaluating derivatives
of $F(z, \tau)$ at $z=0$, whereas we can evaluate derivatives of
$F(z, \tau)$ at $z=1$ to instead recover the moments of the
distribution. The point here is that when the dust settles we
will evaluate $z$ to be less than or equal to one.
Furthermore, we know that the parameter of the geometric distribution $\theta$
must be strictly between zero and one. With these two facts we can safely state that
$| z (1 - \theta) | < 1$. Defining $n \equiv m - m'$ we rewrite the last sum in
Eq.~\ref{eq:si_generating_02} as
\begin{align}
\sum_{m=m'}^{\infty} \left[z (1-\theta)\right]^{m} &= 
\sum_{n=0}^{\infty} \left[z (1-\theta)\right]^{n + m'} \\
&= \left[ z (1 - \theta) \right]^{m'} 
\sum_{n=0}^{\infty} \left[ z (1 - \theta) \right]^{n} \\
&= \left[ z (1 - \theta) \right]^{m'} 
\left( {1 \over 1 - z (1 - \theta)} \right),
\end{align}
where we use the geometric series since, as stated before, $| z (1 - \theta) | <
1$. Putting these results together, the PDE for the generating function is
\begin{equation}
{\partial F \over \partial \tau} = 
{\partial F \over\partial z}
- z {\partial F \over \partial z} - \lambda F
+ \frac{\lambda\theta F}{1-z(1-\theta)}.
\end{equation}
Changing variables to $\xi=1-\theta$ and simplifying gives
\begin{align}
{\partial F \over \partial \tau} + (z - 1) {\partial F \over \partial z} = 
\frac{(z-1)\xi}{1-z\xi}\lambda F.
\label{eq:1state_unreg_015}
\end{align}

@@

To get at the mRNA distribution at steady state we first must solve
Eq.~\ref{eq:1state_unreg_015} setting the time derivative to zero. At
steady-state, the PDE reduces to the ODE
\begin{align}
\deriv[F]{z} = \frac{\xi}{1-z\xi}\lambda F,
\end{align}
which we can integrate as
\begin{align}
\int \frac{dF}{F} = \int \frac{\lambda\xi dz}{1-\xi z}.
\end{align}
The initial conditions for generating functions can be subtle and confusing. The
key fact follows from the definition
$F(z,t) = \sum_m z^m p(m,t)$.
Clearly normalization of the distribution requires that
$F(z=1, t) = \sum_m p(m,t) = 1$.
A subtlety is that sometimes the generating function may be undefined
\textit{at} $z=1$, in which case the limit as $z$ approaches $1$ from
below suffices to define the normalization condition.
We also warn the reader that, while it is frequently convenient to change
variables from $z$ to a different independent variable, one must
carefully track how the normalization condition transforms.

Continuing on, we evaluate the integrals (producing a constant $c$) which gives
\begin{align}
\ln F &= -\lambda \ln(1-\xi z) + c
\\
F &= \frac{c}{(1-\xi z)^\lambda}.
\end{align}
Only one choice for $c$ can satisfy initial conditions, producing
\begin{align}
F(z) = \left(\frac{1-\xi}{1-\xi z}\right)^\lambda
        = \left(\frac{\theta}{1 - z(1-\theta)}\right)^\lambda,
\label{eq:gen_fn}
\end{align}

@@

Let us move from a one-state promoter to a two-state promoter, where one state
has repressor bound and the other produces transcriptional bursts as above.
A schematic of this model is shown as model 5 in
Figure~\ref{fig1:means_cartoons}(C). Although now we have an equation for each
promoter state, otherwise the master equation reads similarly to the one-state
case, except with additional terms corresponding to transitions between promoter
states, namely
\begin{align}
{d\over dt}p_R(m,t) =& 
k_R^+ p_A(m,t) - k_R^- p_R(m,t)
        + (m+1)\gamma p_R(m+1,t) - m\gamma p_R(m,t)
\\
\begin{split}
{d\over dt}p_A(m,t) =& - k_R^+ p_A(m,t) + k_R^- p_R(m,t)
        + (m+1)\gamma p_A(m+1,t) - m\gamma p_A(m,t) 
\\
&- k_i p_A(m,t) + k_i \sum_{m^\prime=0}^m \theta(1-\theta)^{m-m^\prime} p_A(m^\prime,t),
\end{split}
\end{align}
where $p_R(m,t)$ is the probability of the system having $m$ mRNA copies and
having repressor bound to the promoter at time $t$, and $p_A$ is an analogous
probability to find the promoter without repressor bound. $k_R+$ and $k_R^-$
are, respectively, the rates at which repressors bind and unbind to and from the
promoter, and $\gamma$ is the mRNA degradation rate. $k_i$ is the rate at which
bursts initiate, and as before, the geometric distribution of burst sizes has
mean $b=(1-\theta)/\theta$.

@@

@@

f_R(w) = C^{(1)} {_2F_1}(\alpha, \beta, \delta; w)
+ C^{(2)} w^{1-\delta}{_2F_1}(1+\alpha-\delta, 1+\beta-\delta, 2-\delta; w)

@@

\delta &= 1 + k_R^- + k_R^+

@@

and constants $C^{(1)}$ and $C^{(2)}$ to be set by boundary conditions. Solving for

@@

@@

\delta &= 1 + k_R^- + k_R^+.

@@

we can immediately set $C^{(2)}=0$ to discard the second solution.

@@

so $\delta>1$ and therefore $w^{1-\delta}$ blows up as $w\rightarrow0$.

@@

Plugging our solution for $f_R(w) = C^{(1)}{_2F_1}(\alpha, \beta, \delta; w)$

@@

\pderiv[f_R]{w} = C^{(1)}\frac{\alpha\beta}{\delta}
                {_2F_1}(\alpha+1, \beta+1, \delta+1; w),

@@

f_A(w) = \frac{C^{(1)}}{k_R^+}

@@

k_R^- {_2F_1}(\alpha, \beta, \delta; w)
+ w\frac{\alpha\beta}{\delta} {_2F_1}(\alpha+1, \beta+1, \delta+1; w)

@@

f(w) = C^{(1)}\left(1 + \frac{k_R^-}{k_R^+}\right)
        {_2F_1}(\alpha, \beta, \delta; w)
+ w \frac{C^{(1)}}{k_R^+} \frac{\alpha\beta}{\delta}
        {_2F_1}(\alpha+1, \beta+1, \delta+1; w).

@@

w\frac{\alpha\beta}{\delta} {_2F_1}(\alpha+1, \beta+1, \delta+1; w)

@@

(\delta-1)\left(
{_2F_1}(\alpha, \beta, \delta-1; w) - {_2F_1}(\alpha, \beta, \delta; w)

@@

f(w) = \frac{C^{(1)}}{k_R^+}

@@

        {_2F_1}(\alpha, \beta, \delta; w)
+ (\delta-1)\left(
        {_2F_1}(\alpha, \beta, \delta-1; w) - {_2F_1}(\alpha, \beta, \delta; w)

@@

and since $\delta-1 = k_R^+ + k_R^-$, the first and third terms cancel,

@@

f(w) = C^{(1)}\frac{k_R^+ + k_R^-}{k_R^+} {_2F_1}(\alpha, \beta, \delta-1; w).

@@

${_2F_1}(\alpha, \beta, \delta-1; 0) = 1$, so we must have
$C^{(1)} = k_R^+ / (k_R^+ + k_R^-)$ and consequently

@@

$p(m)$ and similarly for $\beta$ and $\delta$. These resemble ratios of
factorials, but since $\alpha$, $\beta$, and $\delta$ are not necessarily

@@

@@

The hypergeometric function is defined by its Taylor series as
\begin{align}
{_2F_1}(a,b,c;z) 
= \sum_{l=0}^\infty
\frac{\Gamma(a + l)\Gamma(b + l)\Gamma(c)}
        {\Gamma(a)\Gamma(b)\Gamma(c + l)}
\frac{z^l}{l!}
\end{align}
for $|z|<1$, and by analytic continuation elsewhere.
If $z\lesssim1/2$ and $\alpha$ and $\beta$ are not too large
(absolute value below 20 or 30),
then the series converges quickly and an accurate numerical representation is
easily computed by truncating the series after a reasonable number of terms.
Unfortunately, we need to evaluate ${_2F_1}$ over mRNA copy numbers fully out
to the tail of the distribution, which can easily reach 50, possibly 100.
From~\eq{eq:p_m_bursty+rep_appdx}, this means evaluating ${_2F_1}$
repeatedly for values of $a$, $b$, and $c$ spanning the full range
from $\mathcal{O}(1)$ to $\mathcal{O}(10^2)$,
even if $\alpha$, $\beta$, and $\delta$
in~\eq{eq:p_m_bursty+rep_appdx} are small,
with the situation even worse if they are not small.
A naive numerical evaluation of the series definition will be
prone to overflow and, if any of $a,b,c<0$, then some successive terms in the
series have alternating signs which can lead to catastrophic cancellations.

@@

This provides a convenient solution for our problem: we need repeated
evaluations of ${_2F_1}(a+m, b+m, c+m; z)$
for fixed $a,b$, and $c$ and many integer values of $m$.
They idea is that we can use arbitrary precision arithmetic to evaluate
${_2F_1}$ for just two particular values of $m$ and then generate
${_2F_1}$ for the other 50-100 values of $m$ using the recurrence
relation.
In fact there are even more sophisticated ways of utilizing the recurrence
relations that might have netted another factor of 2 speed-up, and
possibly as much as a factor of 10, but the method described here had
already reduced the computation time to an acceptable
$\mathcal{O}(\text{1 min})$, so these more sophisticated approaches did
not seem worth the time to pursue.

However, there are two further wrinkles.
The first is that
a naive application of the recurrence relation is numerically unstable.
Roughly, this is because the three term recurrence relations,
like second order ODEs, admit two linearly independent solutions.
In a certain eigenbasis, one of these solutions dominates the other
as $n\rightarrow\infty$, and as $n\rightarrow-\infty$,
the dominance is reversed.
If we fail to work in this eigenbasis, our solution of the recurrence relation
will be a mixture of these solutions and rapidly accumulate numerical error.
For our purposes, it suffices to know that the authors of~\cite{Gil2007}
derived the numerically stable solutions (so-called \textit{minimal solutions})
for several possible choices of $\epsilon_i, \epsilon_j, \epsilon_k$.
Running the recurrence in the proper direction using a minimal solution
is numerically robust and can be done entirely in floating point arithmetic, 
so that we only need to evaluate ${_2F_1}$ with arbitrary precision arithmetic
to generate the seed values for the recursion.

@@

\subsubsection{Particulars}
Let us now demonstrate these generalities for our problem of interest.
In order to evaluate the probability distribution of our
model,~\eq{eq:p_m_bursty+rep_appdx}, we need to evaluate hypergeometric functions
of the form ${_2F_1}(\alpha+m, \beta+m, \delta+m; -b)$
for values of $m$ ranging from $0$ to $\mathcal{O}(100)$.
The authors of~\cite{Gil2007} did not derive a recursion relation
for precisely this case. We could follow their methods and do so ourselves,
but it is much easier to convert to a case that they did consider.
The strategy is to look through the minimal solutions tabulated
in~\cite{Gil2007} and search for a transformation we could apply to
${_2F_1}(\alpha+m, \beta+m, \delta+m; -b)$ that would place the $m$'s
(the variable being incremented by the recursion)
in the same arguments of ${_2F_1}$ as the minimal solution.
After some ``guess and check,'' we found that the transformation
\begin{align}
{_2F_1}(\alpha+m, \beta+m, \delta+m; -b)
=
(1+b)^{-\alpha-m}
        {_2F_1}\left(\alpha+m, \delta-\beta, \delta+m; \frac{b}{1+b}\right),
\label{eq:rec_euler_pretransform}
\end{align}
produces a ${_2F_1}$ on the right hand side that closely resembles
the minimal solutions $y_{3,m}$ and $y_{4,m}$ in Eq.~4.3 in~\cite{Gil2007}.
Explicitly, these solutions are
\begin{align}
y_{3,m}
&\propto
{_2F_1}\left(-\alpha^\prime + \delta^\prime - m,
                -\beta^\prime + \delta^\prime,
                1-\alpha^\prime-\beta^\prime+\delta^\prime-m;
                1-z\right)
\\
y_{4,m}
&\propto
{_2F_1}\left(\alpha^\prime + m,
                \beta^\prime,
                1+\alpha^\prime+\beta^\prime-\delta^\prime+m;
                1-z\right),
\label{eq:minimal_soln_sans_prefac}
\end{align}
where we have omitted prefactors which are unimportant for now.
Which of these two we should use depends on what values $z$ takes on.
Equating $1-z=b/(1+b)$ gives $z=1/(1+b)$, and since $b$ is strictly positive,
$z$ is bounded between 0 and 1.
From Eq.~4.5 in~\cite{Gil2007}, $y_{4,m}$ is the minimal solution
for real $z$ satisfying $0<z<2$, so this is the only minimal solution we need.
